Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2024-12-03 manuscript.Rmd
uppdate statistical section in methods

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot, Marguerite Cogné, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"

abstract: |
Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural land and diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste, namely cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. Our results show that these traditional AFS can store large amounts of carbon, with the average being 156\ Mg\ C\ ha$^{-1}$. The AFS with the highest carbon stocks (forest gardens) stored an average of 210\ Mg\ C\ ha$^{-1}$, close to the values of old-growth tropical forests. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFS in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste’s climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

  journal: "Agroforestry systems"
date: ""

output:
  officedown::rdocx_document: 
    base_format: "bookdown::word_document2"
    reference_docx: misc/word-styles-ref-02.docx
bibliography: misc/references.bib
nocite: |
@Nair1993, @Nair2021, @Orsini1976
csl: misc/springer-basic-author-date.csl
keywords: "Traditional agroforestry, carbon stocks, soil, biomass"
---

Camille Piponiot(*) - Marguerite Cogné - Vincent Freycon - Régis Peltier

@@

(*) email: camille.piponiot-laroche@cirad.fr

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022;@Gibbs2010;@Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO~2~e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to grow as the world’s population increases [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food systems at risk [@Lawrence2015]. There is therefore an urgent need to promote tropical agricultural landscapes that are more resilient to climate change while maintaining high productivity and ecological value [@Harvey2014]. In particular, increasing the amount of carbon stored in the soil and vegetation of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of food production systems [@Lal2004;@Minasny2017].

Agroforestry has attracted attention in recent decades as a set of effective agricultural practices that could contribute to climate change mitigation and adaptation [@Albrecht2003;@Verchot2007;@Lasco2014;@Cardinael2021]. Agroforestry is a collective name for land use systems and technologies where woody perennials (trees, shrubs, bamboos, creepers, etc.) are deliberately used on the same land management units as agricultural crops and/or animals, in some form of spatial arrangement or temporal sequence [@Lundgren1982]. Agroforestry allows for the diversification of agricultural production, making agricultural systems more resilient [@Duffy2021;@Terasaki2023].

Agroforestry systems (AFS) have the potential to store large amounts of carbon in the vegetation and soil, and this carbon storage capacity depends on the characteristics of the AFS considered [@Albrecht2003;@Feliciano2018]. As trees can store large amounts of carbon, the carbon stored in the vegetation of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) has been relatively little studied [@Beillouin2021], is more variable and depends on several factors. Although AFS appear to increase SOC stocks overall, some AFS such as woodlots (i.e., planting trees with intercropping during the establishment phase) may decrease SOC stocks compared to non-AFS, while AFS such as silvopastures have been shown to strongly increase SOC stocks [@Feliciano2018]. Other factors such as climate and previous land use type can also influence SOC in AFS, which may explain the variability of observations reported in the literature [@Feliciano2018;@Ma2020;@Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS could only be evidenced for the topsoil layer [< 10 cm; @Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods. However, population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime are threatening food security [@WB2009;@Molyneux2012]. In this context, productive, resilient, and sustainable agriculture is critical for the country, and agroforestry has been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022;@Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022;@Cogne2024] and can be key elements for biodiversity conservation [@Perfecto2008;@Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. A precise quantification of these carbon stocks would improve understanding of the ecological importance of these AFS at both local and global scales, and help integrate them into national and international climate change mitigation initiatives.

In this study, we aim to better understand traditional AFS in Timor-Leste by quantifying the carbon storage in both biomass and soil of five traditional AFS: cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. We hypothesize that there are measurable differences in aboveground carbon (AGC; H1a) and SOC (H1b) stocks between these different types of traditional AFS. In line with Ma et al. (2020), we anticipate that there is a positive relationship between AGC and SOC stocks in the topsoil layer (0-10 cm; H2a), but no relationship with SOC below 10 cm (H2b).

# Materials and methods

@@

site_summary = site_env[, .(
  coord = paste(round(mean(long),2), round(mean(lat),2), sep = ", "), 
  prec = round(mean(prec)), 
  elev = paste(round(quantile(elev, c(0.025, 0.975))/10)*10, collapse = " and ")),
  .(site)]
```

The research was carried out in the center of the Baucau municipality, located on the north-eastern coast of Timor-Leste. Two communities, or *suco* (a group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases. The first, Gariuai *suco* (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`), is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging between `r site_summary[site=="Gariuai", elev]` m above sea level (asl). The second, the village of Osso-Luga in Samalari *suco* (long-lat coordinates: `r site_summary[site=="Osso-Luga", coord]`), is located on the Matebian foothills based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging between `r site_summary[site=="Osso-Luga", elev]` m asl (Figure\ \@ref(fig:map-sites)). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differ between the two sites (Table\ S\@ref(tab:site-soil)); Gariuai has a clay loam texture whereas Osso-Luga has a clay texture. The two villages also have different recent histories. Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. The AFS in Osso-Luga lay fallow during the period of abandonment.

@@

We used the typology described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair et al. (1993, 2021) and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFS identified were as follows: Crop systems including a wooded Fallow phase (CF), Silvopastures (SP), Young Agroforests (YA), Home Gardens (HG), and Forest Gardens (FG). CF are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts, and squash) which are left fallow after one to three years. SP are vast areas (generally 200 to 500 ha) where cattle are grazed under the shade of trees. YA are agricultural plots where farmers plant rows of trees spaced approximately 5 x 10 m apart, which they maintain for two to ten years. HG are plots of land located close to houses where crops and livestock (poultry, pigs, cattle, etc.) are kept in the shade of numerous trees and palms for multiple uses (wood, fruit, fiber, etc.). FG are generally former HG (over 50 years old) which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

The AFS are presented in more detail in the supplemental information section (Table\ S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

We randomly selected three plots from each of the five traditional AFS identified in each of the two sites (30 plots in total; Figure\ \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06 and 1.25 ha) for all AFS except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

@@

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the center of the sampling plot (Figure\ S\@ref(fig:design)). Within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were taken with a dendrometer.

@@

In each of the 30 sample plots, a composite of three samples was taken from three of the five 10 x 10 m subplots, following a transect covering the intra-plot variability and at a minimum distance of 1 m from any tree (Figure\ S\@ref(fig:design)). Soils were sampled with a 15 cm auger at 0-10 cm and 20-30 cm depth; due to the size of the auger we could not accurately sample the intermediate soil layer. Soils samples were then air-dried and sieved at 2 mm. They underwent standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table\ S\@ref(tab:site-soil)). Total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

In addition, at one location at the center of the plot, soils were sampled with a cylinder of 250 cm^3^ at depths of 2-7 cm, 12-17 cm and 22-27 cm. Each sample was dried at 105°C in a classic oven until a stable dry mass was reached and then sieved to 2 mm. The fine fraction (< 2 mm) was weighed to be used in the SOC stocks calculation (equation \@ref(eq:soc)).

@@

SOC stocks (in Mg\ C\ ha$^{-1}$) in plot $p$ were calculated as follows:

@@

Where $Corg_i$ is the SOC content at depth $i$ (either 0-10, 10-20 or 20-30 cm); $mfine_i$ is the mass of the fine fraction of the soil in the sampling cylinder at depth $i$ (in g), converted into Mg with a factor of 10$^{-6}$; $Splot$ is the surface of a 1-ha plot (10$^8$ cm$^2$); $\Delta d$ is the difference in sample depths (= 10 cm) and $Vcyl$ is the volume of the sampling cylinder (250 cm$^3$). This calculation corresponds to the method recommended by @Poeplau2017 (M4), which correctly takes into account the coarse fraction of the soil. At depth $i$ = 10-20 cm, where SOC content was not measured (see "Soil measurements" section), we estimated SOC content as a function of SOC content at 0-10 cm and 20-30 cm, assuming a negative exponential model of SOC content decrease with soil depth [@Mishra2009;@Murphy2019]:

@@

### Statistical tests

The estimated AGC and SOC stocks were log-transformed for all statistical tests because, as they only take positive values, we expected them to be log-normally distributed. We removed one CF plot at Gariuai from the following statistical tests, as it did not contain any trees exceeding the minimum measurement size (CBH = 30 cm) and its AGC stocks were therefore estimated at 0\ Mg\ C\ ha$^{-1}$ (even if there were trees < 30 cm CBH) and could not be log-transformed. After testing for normality (Shapiro-Wilks test) and homoscedasticity (Levene’s test), we performed a two-way ANOVA on the log-transformed estimates of AGC and SOC stocks, grouped by AFS and site. To test the relationship between log-transformed SOC and AGC stocks, we fitted two linear mixed models. We predicted SOC stocks at two different depths (either 0-10 or 0-30 cm) as a function of AGC stocks (fixed effect) and site (Samalari or Osso-Luga; random effect).

@@

The uncertainty on plot-level AGC was calculated with the AGBmonteCarlo function from the R package BIOMASS, using a (rather conservative) coefficient of variation of 15% on height measurements [@Hunter2013] and large and small errors for 5 and 95% of trees, respectively [@Chave2004;@Rejou-Mechain2017]. To propagate uncertainty on SOC estimates, we used a coefficient of variation of 12% for organic carbon (as recommended by Cirad laboratories in Montpellier, France), and 10% for the mass of the fine fraction [@Page1999].

@@

@@

Total carbon stocks ranged between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks were the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest total carbon stocks were the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks ranged between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table\ \@ref(tab:anova)), but not between sites (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table\ \@ref(tab:anova)). The AFS with the lowest AGC stocks were the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest AGC stocks were the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

It is also interesting to note that, in line with previous results, the age of the trees in the AFS largely explains the carbon stored in their biomass [@Ma2020]. The AFS described in this study reflect different stages of transition in land use, which follow the growth of their trees (Figure\ S\@ref(fig:seq-diagram)). For example, a farmer might plant trees in a CF field, which will then become a YA that can still produce annual crops for the first 2 to 4 years, before the shade from the trees becomes too great. The trees are then used as property markers and, as they mature, produce wood and fruits; the YA then becomes an HG or FG. Similarly, our results show that AGC stocks increase from CF to YA, HG and FG (Table\ \@ref(tab:summary)).

>>>>

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot, Marguerite Cogné, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"
abstract: |
Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and/or other woody perennials into agricultural land thereby  diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste, namely cropping systems with fallow (CF), silvopastures (SP), young agroforests (YA), home gardens (HG), and forest gardens (FG). Our results show that these traditional AFS can store large amounts of carbon, with the average being 155\ Mg\ C\ ha$^{-1}$. The AFS with the highest carbon stocks (FG) stored an average of 210\ Mg\ C\ ha$^{-1}$,  close to the values of old-growth tropical forests. Other AFS stored on average: 107\ Mg\ C\ ha$^{-1}$ in CF, 158\ Mg\ C\ ha$^{-1}$ in SP, 132\ Mg\ C\ ha$^{-1}$ in YA, and 167\ Mg\ C\ ha$^{-1}$ in HG.  Biomass carbon stocks varied significantly between AFS (which differed in tree cover), while soil carbon stocks were less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFS in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste’s climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

  journal: "Agroforestry systems"
date: ""
output:
  officedown::rdocx_document: 
    base_format: "bookdown::word_document2"
    reference_docx: misc/word-styles-ref-02.docx
bibliography: misc/references.bib
nocite: |
csl: misc/springer-basic-author-date.csl
keywords: "Traditional agroforestry, carbon stocks, soil, biomass"
editor_options: 
  markdown: 
    wrap: 72
---

Camille Piponiot(\*) - Marguerite Cogné - Vincent Freycon - Régis Peltier

@@

(\*) email: [camille.piponiot-laroche\@cirad.fr](mailto:camille.piponiot-laroche@cirad.fr){.email}

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022; @Gibbs2010; @Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO~2~e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to grow as the world’s population increases [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food production systems at risk [@Lawrence2015]. Therefore, there is an urgent need to promote tropical agricultural landscapes that are more resilient to climate change while maintaining high productivity and ecological value [@Harvey2014]. Increasing the amount of carbon stored in the soil and vegetation biomass of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of food production systems [@Lal2004; @Minasny2017].

Agroforestry has attracted attention in recent decades as a set of effective agricultural practices that could contribute to climate change mitigation and adaptation [@Albrecht2003; @Verchot2007; @Lasco2014; @Cardinael2021]. Agroforestry is a collective name for land use systems and technologies where woody perennials (trees, shrubs, bamboos, creepers, etc.) are deliberately used on the same land management units as agricultural crops and/or animals, in some form of spatial arrangement or temporal sequence [@Lundgren1982]. Therefore, agroforestry practices allow for the diversification of agricultural production, making agricultural systems more resilient to climate change [@Duffy2021; @Terasaki2023].

Agroforestry systems (AFS) have the potential to store large amounts of carbon in the vegetation biomass and soil, and this carbon storage capacity depends on the type of the AFS in addition to environmental and edaphic conditions [@Albrecht2003; @Feliciano2018]. Since trees can store large amounts of carbon, the carbon stored in vegetation aboveground biomass of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) is more variable: although AFS appear to increase SOC stocks overall in tropical regions, some AFS such as woodlots (i.e., planting trees with intercropping during the establishment phase) may decrease SOC stocks compared to non-AFS, while AFS such as silvopastures have been shown to strongly increase SOC stocks [@Feliciano2018]. Other factors such as climate and previous land use type can also influence SOC in AFS, which may explain the variability of observations reported in the literature [@Feliciano2018; @Ma2020; @Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS was only evidenced to a depth of 10 cm [@Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods. However, population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime are threatening food security [@WB2009; @Molyneux2012]. In this context, productive, resilient, and sustainable agriculture is critical for the country and the livelihood of its people. To address this, AFS have been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022; @Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022; @Cogne2024] and can be key elements for biodiversity conservation [@Perfecto2008; @Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. A precise quantification of these carbon stocks will improve understanding of the ecological importance of AFS at both local and global scales, and help integrate them into national and international climate change mitigation initiatives.

In this study, we aim to better understand traditional AFS in Timor-Leste by quantifying the carbon storage in both vegetation biomass and soil of five traditional AFS: cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. We hypothesize that there are measurable differences in carbon in aboveground vegetation biomass (hereinafter referred to as above ground carbon or AGC; Hypothesis 1a) and SOC (Hypothesis 1b) stocks between these different types of traditional AFS. In line with @Ma2020, we anticipate that there is a positive relationship between AGC and SOC stocks in the topsoil layer (0-10 cm; Hypothesis 2a), but no relationship with SOC below 10 cm (Hypothesis 2b).

# Materials and Methods

@@

site_summary = site_env[, .(
  coord = paste(round(mean(lat),2), round(mean(long),2), sep = ", "), 
  prec = round(mean(prec)), 
  elev = paste(round(quantile(elev, c(0.025, 0.975))/10)*10, collapse = " and ")),
  .(site)]
```

The research was carried out in the center of the Baucau municipality, located on the north-eastern coast of Timor-Leste. Two communities, or *suco* (a group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases. The first, Gariuai *suco* (Lat-Lon coordinates: `r site_summary[site=="Gariuai", coord]`), is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging between `r site_summary[site=="Gariuai", elev]` m above sea level (masl). The second, the village of Osso-Luga in Samalari *suco* (Lat-Lon coordinates: `r site_summary[site=="Osso-Luga", coord]`), is located on the Matebian foothills based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging between `r site_summary[site=="Osso-Luga", elev]` masl (Figure \@ref(fig:map-sites)). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differ between the two sites (Table S\@ref(tab:site-soil)); Gariuai has a clay loam texture whereas Osso-Luga has a clay texture. The two villages also have different histories. Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. The AFS in Osso-Luga lay fallow during the period of abandonment. The land use of these 2 villages is typical of the Baucau district. Land use in these two villages is typical of the Baucau district. The 5 AFS described in this article [see detailed description in @Cogne2024] are found in the valleys and high plateaux, while non-AFS such as rainfed rice and horticultural plots are more likely to be found in the lowlands. Other non-agricultural land uses are also present, such as limestone and limestone quarrying. Some natural forest parks are also preserved around mountain peaks, such as Matebian or Mundo Perdido. Villages are concentrated near paved roads and on hilltops, more rarely near rivers because of the risk of flooding.

@@

We used the typology of AFS described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair *et al.* (1993, 2021) and adapted to the local context with semi-structured interviews previously conducted in another study at the same sites [@Cogne2024]. The five AFS identified were as follows: Crop systems including a wooded Fallow phase (CF), Silvopastures (SP), Young Agroforests (YA), Home Gardens (HG), and Forest Gardens (FG). The AFS are presented in more detail in the supplemental information section (Table S\@ref(tab:afs-typology)) and in @Cogne2024.

Briefly, the CF are plantations of a few non-perennial crops (e.g., maize [*Zea mays*], sweet potatoes [*Ipomoea batatas*], groundnuts [*Arachis hypogaea*], and squash gourd [*Cucurbita moschata*]) which are left fallow after one to three years of cropping. SP are vast areas (generally 200 to 500 ha) where cattle (*Bos taurus*, *Bos domesticus*, *Bubalus bubalis*) are grazed under the shade of palms, like coconut (*Cocos nucifera*), and trees, like tamarind (*Tamarindus indica*) or eucalyptus (*Eucalyptus alba*) and jujube (*Ziziphus sp.*). YA are agricultural plots where farmers plant rows of trees, like teak (*Tectona grandis*) and mahogany (*Swietenia mahagoni*), spaced approximately 5 x 10 m apart, which are maintained for two to ten years and then either cut for wood or left to grow to become HG or FG (see Figure S\@ref(fig:seq-diagram)). HG are plots of land located close to houses where crops and livestock (poultry [*Gallus gallus domesticus*], pigs [*Sus domesticus*], cattle, etc.) are kept in the shade of numerous trees, like Breadfruit (*Artocarpus altilis*), candlenut (*Aleurites moluccanus*) and palms, like areca palm (*Areca catechu*), for multiple uses (wood, fruit, fiber, etc.). FG are generally former HG (over 50 years old) which have been abandoned, for example as a result of the war, and are now used mainly for gathering fruits (e.g. mango [*Mangifera indica*]) and non-timber forest products (e.g. candlenut) with limited human influence.

@@

We randomly selected three plots from each of the five traditional AFS identified in each of the two sites (30 plots in total; Figure \@ref(fig:map-sites)). Sampling plots were defined as the entire agricultural plot (area between 0.06 and 1.25 ha) for all AFS except SP plots, which were generally much larger than 1 ha. In SP plots, a 1 ha sampling plot was delineated within the agricultural plot. Tree and soil measurements (described below) were carried out between June and August 2021.

@@

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the center of the sampling plot (Figure S\@ref(fig:design)). Within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were taken with a dendrometer.

@@

In each of the 30 sample plots, a composite of three samples was taken from three of the five 10 x 10 m subplots, following a transect covering the intra-plot variability and at a minimum distance of 1 m from any tree (Figure S\@ref(fig:design)). Soils were sampled with a 15 cm auger at 0-10 cm and 20-30 cm depth; due to the size of the auger we could not accurately sample the intermediate soil layer. Soils samples were then air-dried and sieved at 2 mm. They underwent standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table S\@ref(tab:site-soil)). Total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

In addition, at one location at the center of the plot, soils were sampled with a cylinder of 250 cm^3^ at depths of 2-7 cm, 12-17 cm and 22-27 cm. Each sample was dried at 105°C in a classic oven until a stable dry mass was reached and then sieved to 2 mm. The fine fraction (\< 2 mm) was weighed to be used in the SOC stocks calculation (equation \@ref(eq:soc)).

@@

SOC stocks (in Mg C ha$^{-1}$) in plot $p$ were calculated as follows:

@@

Where $Corg_i$ is the SOC content at depth $i$ (either 0-10, 10-20 or 20-30 cm); $mfine_i$ is the mass of the fine fraction of the soil in the sampling cylinder at depth $i$ (in g), converted into Mg with a factor of 10$^{-6}$; $Splot$ is the surface of a 1-ha plot (10$^8$ cm$^2$); $\Delta d$ is the difference in sample depths (= 10 cm) and $Vcyl$ is the volume of the sampling cylinder (250 cm$^3$). This calculation corresponds to the method recommended by @Poeplau2017 (M4), which correctly takes into account the coarse fraction of the soil. At depth $i$ = 10-20 cm, where SOC content was not measured (see "Soil measurements" section), we estimated SOC content as a function of SOC content at 0-10 cm and 20-30 cm, assuming a negative exponential model of SOC content decrease with soil depth [@Mishra2009; @Murphy2019]:

@@

@@

The uncertainty on plot-level AGC was calculated with the AGBmonteCarlo function from the R package BIOMASS, using a (rather conservative) coefficient of variation of 15% on height measurements [@Hunter2013] and large and small errors for 5 and 95% of trees, respectively [@Chave2004; @Rejou-Mechain2017]. To propagate uncertainty on SOC estimates, we used a coefficient of variation of 12% for organic carbon (as recommended by Cirad laboratories in Montpellier, France), and 10% for the mass of the fine fraction [@Page1999].

@@

### Model calibration

We calibrated the following models for AGC stocks of plot $i$ in site $k$ and AFS type $j$:

$$AGC_{i,j,k}  = \mu_A + \alpha_{A,j} + \beta_{A,k} + \epsilon_{A,i}$$
with $\mu_A$ the average AGC stocks in our study area, $\alpha_{A,j}$ and  $\beta_{A,k}$ the random AFS and site effects, respectively distributed as:

$$\alpha_{A,j} \sim \mathcal{N}(0,\sigma_{\alpha A}^2)$$

$$\beta_{A,k} \sim \mathcal{N}(0,\sigma_{\beta A}^2)$$
And $\epsilon_{A,i}$ the random error, distributed as:

$$\epsilon_{A,i} \sim \mathcal{N}(0,\sigma_A^2)$$
$\sigma_A$, $\sigma_{\alpha A}$, and $\sigma_{\beta A}$ are the standard deviation of the random error, the AFS effect and the site effect respectively.

Similarly, we modelled SOC stocks of plot $i$ in site $k$ and AFS type $j$ as:

$$SOC_{i,j,k} = \mu_S +  \alpha_{Sj} + \beta_{Sk} + \theta_1 agc_i + \theta_2 fine_i+ \epsilon_{S,i}$$
with $\mu_S$ the average SOC stocks in our study area; $\theta_1$ the effect of AGC stocks on SOC stocks and $agc_i$ the standardised AGC stocks of plot $i$; $\theta_2$ the effect of fine soil fraction (clay + fine silt) on SOC stocks and $fine_i$ the standardised fine soil fraction of plot $i$. We standardised AGC stocks and fine soil fraction by scaling and centring values.

$\alpha_{S,j}$ and $\beta_{S,k}$ are the random AFS and site effects, respectively distributed as:

$$\alpha_{S,j} \sim \mathcal{N}(0,\sigma_{\alpha S}^2)$$

$$\beta_{S,k} \sim \mathcal{N}(0,\sigma_{\beta S}^2)$$
And $\epsilon_{S,i}$ the random error, distributed as:

$$\epsilon_{S,i} \sim \mathcal{N}(0,\sigma_S^2)$$
$\sigma_S$, $\sigma_{\alpha S}$, and $\sigma_{\beta S}$ are the standard deviation of the random error, the AFS effect and the site effect respectively.

Calibration was carried out using an adapted form of the Hamiltonian Monte Carlo using Stan’s programming language [@Carpenter2017]. We used the standard half-normal distribution as the prior for all standard deviation parameters, and the standard normal distribution for both $\theta_1$ and $\theta_2$ parameters. To propagate uncertainties on the model outputs, we calibrated the model for each of the 1000 iterations and sampled one set of parameter values in the posterior distributions for each iteration.

@@

Total carbon stocks ranged between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]` Mg C ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]` Mg C ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`% (Table \@ref(tab:summary); Figure \@ref(fig:barplot)). The AFS with the lowest total carbon stocks were the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]` Mg C ha$^{-1}$; the AFS with the highest total carbon stocks were the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]` Mg C ha$^{-1}$ (Table \@ref(tab:summary); Figure \@ref(fig:barplot)).

@@

AGC stocks ranged between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]` Mg C ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]` Mg C ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`% (Table \@ref(tab:summary); Figure \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)), but not between sites (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The AFS with the lowest AGC stocks were the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]` Mg C ha$^{-1}$; the AFS with the highest AGC stocks were the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]` Mg C ha$^{-1}$ (Table \@ref(tab:summary); Figure \@ref(fig:barplot)).

@@

It is also interesting to note that, in line with previous results, the age of the trees in the AFS largely explains the carbon stored in their biomass [@Ma2020]. The AFS described in this study reflect different stages of transition in land use, which follow the growth of their trees (Figure S\@ref(fig:seq-diagram)). For example, a farmer might plant trees in a CF field, which will then become a YA that can still produce annual crops for the first 2 to 4 years, before the shade from the trees becomes too great. The trees are then used as property markers and, as they mature, produce wood and fruits; the YA then becomes an HG or FG. Similarly, our results show that AGC stocks increase from CF to YA, HG and FG (Table \@ref(tab:summary)).
